Macros for straightforward parallelism in Clojure

### 2016-04-19 paper/paper.Rnw
paper

@@

Clojure is a lisp--like language running on the Java Virtual Machine.\cite{Hickey2008}
The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.\cite{Kraus2009}
Because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.

@@

As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.
Clojure builtins such as pmap\footnotemark[1] and reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
Third---party libraries like Tesser\footnotemark[3] and Claypoole\footnotemark[4] provide more data parallel APIs with slightly different goals than the builtin functions.
Developers have a good relationship with data parallel problems\cite{Okur2012a}, but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea, Rodr}).
An astute reader may observe that both Claypoole and the Clojure standard library include task parallel functions.
These will be discussed in the Section \ref{label:related_work}

@@

Using Clojure's macro system, I have implemented a set of macros which allow developers to take advantage of Clojure's parallelism potential when their existing code is written such that parallelism is exposed through control flow.
I have shown that it is possible to attain reasonable degrees of parallelism with minimal code changes, with respect to serial code, using these macros.
The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.

\section{Related Work} \label{label:related_work}
% only discuss task based parallelism in this section
Clojure has access to all of the JVM, so it has access to the Java \code{ForkJoinPool} library\cite{Lea2000}.
The \code{ForkJoinPool} library allows Java programmers to create lightweight, recursive tasks, then execute them on an efficient, work stealing, \code{ExecutionEngine}.
Users of a \code{ForkJoinPool} create a subclass of \code{RecursiveTask} to compute the value of some function.
Each of these tasks may create more \code{RecursiveTask}s, submit them to the \code{ForkJoinPool}, then wait for their subtasks to complete, without blocking the pool.
The execution engine for the \code{ForkJoinPool} is a thread pool.
Each thread maintains it's own queue of tasks to work on.
Whenever a thread runs out of tasks to work on, it steals tasks from other threads.
Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
Clojure programmers can use the \code{ForkJoinPool} from Clojure, but the interface isn't exactly programmer friendly.

Clojure also has some built in support for task parallelism via \code{future}\footnote{\url{https://clojuredocs.org/clojure.core/future}}.
Each call to future creates a thread to compute the result of a function call in the background.
When a user is ready to access a value, they can \code{deref} the future to get the computed value (\code{deref} will block until the value is computed).
For a user, \code{future}s initially seem like a fairly natural way to parallelize recursive functions.
Unfortunately, Futures only work well when a small number of tasks are created or when the tasks do a lot of blocking I/O.
Thread creation overhead is high, and people users must be careful not to create an excessive number of threads or create threads to do too little work.
Claypoole offers many improvements to the builtin Clojure \code{future}s, but it still creates a thread per task.
The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future}s interface.
My code also creates tasks for every thread, but these tasks are created on a Java \code{ForkJoinPool}, so the tasks are much more lightweight.

@@

\subsection{Other languages}
In other languages, frameworks like Cilk++\footnote{\url{https://www.cilkplus.org/}}, OpenMP\footnote{\url{http://openmp.org/wp/}}, Threading Building Blocks\footnote{\url{https://www.threadingbuildingblocks.org/}} all exist.
Some of these require compiler support and advanced dependence analysis to guarantee correctness.
These libraries all have great performance, but they can't be leveraged from Clojure.
The macros I've implemented are an attempt to bring some of the niceties of these libraries and compiler extensions to Clojure, without the need to compiler support or complicated dependence analysis.

\section{Mostly Pure Functions} \label{label:mostly_pure}
Before discussing the macros I've implemented, I need to loosely define a ``mostly pure'' function.
A mostly pure function is a thread---safe function with no side effects directly impacting the values of user-defined values, at the current level of abstraction.
Mostly pure functions can be reordered (or interleaved) without impacting the values of user variables, although the change may impact I/O behavior and output order of a program.
In some cases, the order of certain side effects may not matter to a programmer.
For example, it may not matter if we reorder \mintcode{(download file1)} and \mintcode{(download file2)} for a programmer writing a web scraper, but it may matter to a programmer writing a I/O constrained server.
When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions (the call is ``portable''), we can reorder them subject to these constraints:

@@

Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
In this paper, we rely on the judgment of the programmer to tell us when a mostly pure function is portable.

@@

\section{parlet}
The first of the parallel macros is called \parlet{}.
\parlet{} is a modified version of the Clojure \code{let} form.
The \parlet{} macro has exactly the same behavior as Clojure's \code{let}, but it evaluates all of its bindings in parallel.
For example, suppose I had some long running function \code{foo}.
I need to add the result of two calls to this function.
In Figure \ref{fig:parlet_basic}, we use \parlet{} to make two calls to \code{foo}, then add the results.

\begin{figure}[h]
\centering
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\begin{minted}{clojure}
(parlet
  [a (foo value1)
   b (foo value2)]
   ... ; some other code here
  (+ a b))
\end{minted}
\caption{Example of a parlet form}
\label{fig:parlet_basic}
\end{figure}

In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}\cite{Lea2000}.
The calls to \code{foo} are both forked immediately, then we attempt to evaluate the body of the \parlet{}.
This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.
Additionally, since the \code{ForkJoinPool} is designed for recursive workloads, tasks which are currently executing can create new tasks, submit them to the pool, then wait for the task to complete, without blocking execution.
This means that nested \parlet{} forms (Figure \ref{fig:parlet_noblock}) will not block each other.

\begin{figure}[h]
\centering
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\begin{minted}{clojure}
(parlet
  [a (foo 100)]
  ;; some other code not using a
  (parlet
    [b (foo 200)]
    (+ a b)))
\end{minted}
\caption{Nested parlet forms}
\label{fig:parlet_noblock}
\end{figure}

Code like this may not arise when the code is human generated, but it may arise when the code is generated by another macro.
We will see some examples of this later.

\subsection{Dependencies}
The \parlet{} macro also supports simple dependency detection.
Clojure \code{let} forms can use names defined previously in the let, and the bindings are evaluated from first to last.

@@

Without the \parlet{}, the let form in Figure \ref{fig:parlet_dep} would evaluate to 2.
If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
So, the \parlet{} macro looks through the bindings in the macro to determine if any of them depend on any of the names defined previously in the macro.
If there are any, the macro will halt the compiler and report an error to the user.

@@

% \section{parexpr}

% The \parexpr{} macro breaks down expressions and (aggressively) evaluates them in parallel.
% Suppose again that I had a long running function \code{foo} which I wanted to call twice, and add the results.
% The code in Figure \ref{fig:parexpr} will make both of the long running calls to \code{foo} in parallel.
% The \parexpr{} macro crawls the expression and expands it into multiple nested \parlet{} forms, filling each \parlet{} with as many evaluations as it can.

% \begin{figure}[h]
% \centering
% \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
% \begin{minted}{clojure}
% (parexp (+ (foo value1) (foo value2)))
% \end{minted}
% \caption{Example using parexpr to evaluate long running functions}
% \label{fig:parexpr}
% \end{figure}

@@

For an example see Figure \ref{fig:fib_code}.
This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.

\begin{figure}[h]
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\centering
\begin{minted}{clojure}
(defparfun fib [n] (< n 35)
  (if (or (= 0 n) (= 1 n))
    1
    (+
     (fib (- n 1))
     (fib (- n 2)))))
\end{minted}
\caption{Fibonacci function with \defparfun{} added}
\label{fig:fib_code}
\end{figure}

@@

\begin{figure}[h]
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\centering
\begin{minted}{clojure}
(defun fib [n]
  (if (or (= 0 n) (= 1 n))
    1
    (+
     (fib (- n 1))
     (fib (- n 2)))))
\end{minted}
\caption{Serial Fibonacci function}
\label{fig:fib_serial}
\end{figure}

First we will look at the classical recursive Fibonacci example.
Figure \ref{fig:fib_serial} shows the serial benchmark code; \ref{fig:fib_code} shows the code parallel benchmark code.
In Figure \ref{fig:fibparfun} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
Each benchmark computes \mintcode{(fib 39)}.
We see that we get about a 3x speedup with $6$ cores.
This speedup isn't quite what we would hope to see, but, as can be seen in Figure \ref{fig:javafib}, the handwritten Java \code{ForkJoinPool} implementation gets about the same speedup with $6$ cores on these virtual machines.
Previous \code{ForkJoinPool} benchmarks have shown much better speedups for similar code\cite{Lea2000}, so I suspect that the virtual machine configuration is somewhat responsible for the discrepancy in results.

<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
@

<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
jfib <- plot_means_error_for_benchmark(data, "fib", "fj")
@

The large variance we see in the Clojure benchmarks is somewhat disturbing, especially since it does not show up in the Java results.
Since it does not appear in the Java benchmarks, it is not a result of variable performance in the Google Cloud Platform virtual machines.
In Figure \ref{fig:fib_sd} I show the standard deviation of the mean runtime for the serial and parallel Clojure Fibonacci functions, along with their Java counterparts.
Notice that the Java results do not have nearly as high deviations from mean runtime.
While I cannot completely explain the variability, it seems to be caused by the increased pressure Clojure's function implementation and the \code{ForkJoinPool} wrapper tasks places on the JVM garbage collector.
Every Clojure function is an \code{Object} created following the Clojure \code{IFn} interface\footnote{\url{http://clojure.org/reference/special_forms\#fn}}.
When running on the \code{ForkJoinPool}, each function is further wrapped in a \code{RecursiveTask} object, causing additional allocations.
This effectively moves the stack for the recursive function to the heap.
There is so upside to this; it eliminates the stack depth limit (which is an issue in Clojure because Clojure cannot implement tail call optimization), the large number of allocations creates a large amount garbage collector pressure.
The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
We can avoid excessive task creation by controlling the granularity of parallelism, to an extent.
The Fibonacci example highlights this problem because the function call overhead greatly exceeds the amount of work each call is doing.
We will see an example for which this is not the case in Section \ref{label:id3_bench}.

<<fib_sd, fig.pos="H", fig.cap=" ", echo=F, out.width='\\linewidth', fig.align='center'>>=
fib <- data[data$spec.name %like% "fib.big",]

@@

barplot(t(as.matrix(d[,nnames])), beside=TRUE, names=d$cores, legend=nnames,
        main = "Fibonacci standard deviations", xlab="cores", ylab="mean runtime standard deviation")
@

@@

<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
id3 <- plot_means_error_for_benchmark(data, "id3", "parfun")
@

@@

>>>>

\usepackage{subcaption}

@@

Clojure is a lisp--like language running on the Java Virtual Machine\cite{Hickey2008}.
The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming\cite{Kraus2009}.
Because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.

@@

As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.
Clojure built ins such as \code{pmap}\footnotemark[1] and reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
Third---party libraries like Tesser\footnotemark[3] and Claypoole\footnotemark[4] provide more data parallel APIs with slightly different goals than the builtin functions.
Developers have a good relationship with data parallel problems\cite{Okur2012a}, but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea, Rodr}).

@@

Using Clojure's macro system, I have implemented a set of macros which allow developers to take advantage of Clojure's task parallelism potential.
I have shown that it is possible to attain reasonable degrees of parallelism with minimal modification to existing serial code.

\section{Related Work} \label{label:related_work}
% only discuss task based parallelism in this section
Clojure has access to all of the JVM, so it has access to the Java \code{ForkJoinPool} library\cite{Lea2000}.
The \code{ForkJoinPool} library allows Java programmers to create lightweight, recursive tasks which are executed by a thread pool.
Each thread in the pool maintains a queue of tasks to work on.
Whenever a thread runs out of tasks to work on, it steals tasks from other threads.
Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
Users of a \code{ForkJoinPool} create a subclass of \code{RecursiveTask} to compute the value of some function.
Each of these tasks may create more \code{RecursiveTask}s, submit them to the \code{ForkJoinPool}, then wait for their subtasks to complete, without blocking the pool.
Clojure programmers can use the \code{ForkJoinPool} libraries from Clojure, but the interface isn't exactly programmer friendly (or idiomatic).

Clojure also has built in support for task parallelism via \code{future}\footnote{\url{https://clojuredocs.org/clojure.core/future}}.
Each call to \code{future} spawns a thread to compute the result of a function call in the background.
When a user is ready to access the computed value, they can \code{deref} the future to it.
\code{deref} will block until the value is computed.
For a user, \code{future} initially seem like a fairly natural way to parallelize recursive functions.
Unfortunately, \code{future} only works well when a small number of tasks are created, or when the tasks do a lot of blocking I/O.
Because thread creation overhead is high, users must be careful not to create an excessive number of threads or create threads which do too little work.
Claypoole's implementation of \code{future} offers many improvements to the built in \code{future}.
Their future implementation executes tasks created with \code{future} in a fixed sized thread pool, but, the \code{future} task will block the thread executing it if the task creates a new task, then immediately \code{deref}s it.
This means that a programmer must be very careful not to create tasks in a Claypoole thread pool which depend on other tasks, otherwise their code may deadlock.
The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future}s interface.
My code also creates many tasks, but these tasks are created on a Java \code{ForkJoinPool}, so the tasks are much more lightweight, and they can be composed recursively.

@@

\subsection{Other languages}
In other languages, frameworks like Cilk++\footnote{\url{https://www.cilkplus.org/}}, OpenMP\footnote{\url{http://openmp.org/wp/}}, Threading Building Blocks\footnote{\url{https://www.threadingbuildingblocks.org/}} have been implemented.
Some of these require compiler support and advanced dependence analysis to guarantee correctness.
The macros I've implemented are an attempt to bring some of the niceties of these libraries and compiler extensions to Clojure, without the need to compiler support or complicated dependence analysis.

\section{Mostly Pure Functions} \label{label:mostly_pure}
Before discussing the macros I've implemented, I need to loosely define a ``mostly pure'' function.
A mostly pure function is a thread---safe function with no side effects directly impacting the values of user-defined values, at the current level of abstraction.
Mostly pure functions can be reordered (or interleaved) without impacting the values of user variables, although the change may impact I/O behavior and output order of a program.
In some cases, the order of certain side effects may not matter to a programmer.
For example, it may not matter if we reorder \mintcode{(download file1)} and \mintcode{(download file2)} for a programmer writing a web scraper, but it may matter to a programmer writing a I/O constrained server.
When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions, we can reorder them subject to these constraints:

@@

Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
In this paper, we rely on the judgment of the programmer to tell us when a function can be considered mostly pure.

@@

\begin{figure*}[t!]
\begin{subfigure}[b]{0.5\textwidth}
\centering
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\begin{minted}{clojure}
(parlet
  [a (foo value1)
   b (foo value2)]
  ;; some other code here
  (+ a b))
\end{minted}
\subcaption{Example of a parlet form}
\label{fig:parlet_basic}
\end{subfigure}%
\begin{subfigure}[b]{0.5\textwidth}
\centering
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\begin{minted}{clojure}
(let
  [a (fork (new-task (fn [] (foo 1))))
   b (fork (new-task (fn [] (foo 2))))]
  ;; some other code here
  (+ (join a) (join b)))
\end{minted}
\subcaption{An expanded parlet form}
\label{fig:parlet_basic_expanded}
\end{subfigure}
\caption{}
\label{fig:parlet_example}
\end{figure*}

\newpage

\section{parlet}
The first of the parallel macros is called \parlet{}.
\parlet{} is a modified version of the Clojure \code{let} form.
The \parlet{} macro has exactly the same behavior as Clojure's \code{let}, but it evaluates all of its bindings in parallel.
For example, suppose I had some long running function \code{foo}.
I need to add the result of two calls to this function.
In Figure \ref{fig:parlet_example}, we use \parlet{} to make two calls to \code{foo}, then add the results.

In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{RecursiveTask}s in a \code{ForkJoinPool}.
The calls to \code{foo} are both forked immediately, then we attempt to evaluate the body of the \parlet{}.
Each use of \code{a} and \code{b} is replaced with a call to \code{join}, to get the computed value.
This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.
Additionally, since the \code{ForkJoinPool} is designed for recursive workloads, tasks which are currently executing can create new tasks, submit them to the pool, then wait for the task to complete, without blocking tasks created by other \parlet{} calls.
So, \parlet{} forms can be nested; a programmer does not have to worry if functions called in the bindings of a \parlet{} form also use \parlet{}.

\subsection{Dependencies}
The \parlet{} macro also supports simple dependency detection.
Clojure \code{let} forms can use names defined previously in the same \code{let}.
The bindings are evaluated from first to last.

@@

Without the \parlet{}, the let form in Figure \ref{fig:parlet_dep} would evaluate to 2.
If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
So, the \parlet{} macro will halt the compiler and report an error to the user if any dependencies are found in the \parlet{} form.

@@

@@

For an example see Figure \ref{fig:fib_parfun}.
This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.

\begin{figure*}[t!]
\begin{subfigure}[b]{0.5\textwidth}
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\centering
\begin{minted}{clojure}
(defn fib [n]
  (if (or (= 0 n) (= 1 n))
    1
    (+
     (fib (- n 1))
     (fib (- n 2)))))
\end{minted}
\subcaption{Serial Fibonacci function}
\label{fig:fib_serial}
\end{subfigure}%
\begin{subfigure}[b]{0.5\textwidth}
\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
\centering
\begin{minted}{clojure}
(defparfun fib [n] (< n 35)
  (if (or (= 0 n) (= 1 n))
    1
    (+
     (fib (- n 1))
     (fib (- n 2)))))
\end{minted}
\subcaption{Fibonacci function with \defparfun{} added}
\label{fig:fib_parfun}
\end{subfigure}
\caption{}
\label{ref:fib_code}
\end{figure*}

@@

First we will look at the classical recursive Fibonacci example.
Figure \ref{fig:fib_serial} shows the serial benchmark code; \ref{fig:fib_parfun} shows the code parallel benchmark code.
In Figure \ref{fig:cljfib} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
Each benchmark computes \mintcode{(fib 39)}.
We see that we get about a 3x speedup with $6$ cores.
This speedup isn't quite what we would hope to see, but, as can be seen in Figure \ref{fig:javafib}, the handwritten Java \code{ForkJoinPool} implementation gets about the same speedup with $6$ cores on these virtual machines.
Previous \code{ForkJoinPool} benchmarks have shown much better speedups for similar code\cite{Lea2000}, so I suspect that the virtual machine configuration is somewhat responsible for the discrepancy in results.

<<cljfib, fig.pos="h", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
@

<<javafib, fig.pos="h", fig.cap="Fibonacci Java Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
jfib <- plot_means_error_for_benchmark(data, "fib", "fj")
@

The large variance we see in the Clojure benchmarks is somewhat disturbing, especially since it does not show up in the Java results.
Since it does not appear in the Java benchmarks, it is not a result of variable performance in the Google Cloud Platform virtual machines.
In Figure \ref{fig:fib_sd} I show the standard deviation of the mean runtime for the serial and parallel Clojure Fibonacci functions, along with their Java counterparts.
Notice that the Java results do not have nearly as high deviations from mean runtime.
While I cannot completely explain the variability, it seems to be caused by the increased pressure Clojure's function implementation and the \code{ForkJoinPool} wrapper tasks places on the JVM garbage collector.
Every Clojure function is an \code{Object} created following the Clojure \code{IFn} interface\footnote{\url{http://clojure.org/reference/special_forms\#fn}}.
When running on the \code{ForkJoinPool}, each function is further wrapped in a \code{RecursiveTask} object, causing additional allocations.
This effectively moves the stack for the recursive function to the heap.
There is so upside to this; it eliminates the stack depth limit (which is an issue in Clojure because Clojure cannot implement tail call optimization), but the large number of allocations creates a large amount garbage collector pressure.
The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
We can avoid excessive task creation by controlling the granularity of parallelism, to an extent.
The Fibonacci example highlights this problem because the function call overhead greatly exceeds the amount of work each call is doing.
We will see an example for which this is not the case in Section \ref{label:id3_bench}.

<<fib_sd, fig.pos="h", fig.cap="Fibonacci standard deviations", echo=F, out.width='\\linewidth', fig.align='center'>>=
fib <- data[data$spec.name %like% "fib.big",]

@@

barplot(t(as.matrix(d[,nnames])), beside=TRUE, names=d$cores, legend=nnames,
        xlab="cores", ylab="mean runtime standard deviation") #, main = "Fibonacci standard deviations")
@

@@

<<id3parfun, fig.pos="h", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
id3 <- plot_means_error_for_benchmark(data, "id3", "parfun")
@

@@

The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.
